A finite-state morphological transducer for Kyrgyz

### 2011-10-14 dev/paper/paper.tex
more additions and some fixes

\documentclass[a4paper,12pt,onecolumn,oneside]{article}

\usepackage{polyglossia}
\setdefaultlanguage[variant=australian]{english}

@@

\usepackage[in]{fullpage}

@@

\usepackage[colorlinks=true,citecolor=black,linkcolor=black,urlcolor=blue]{hyperref}

\title{A finite-state morphological transducer for Kyrgyz}
\author{Jonathan North Washington, Mirlan Ipasov, and Francis Tyers\\\href{mailto:jonwashi@indiana.edu}{jonwashi@indiana.edu}, \href{mailto:mipasov@gmail.com}{mipasov@gmail.com}, \href{mailto:ftyers@prompsit.com}{ftyers@prompsit.com}}
\date{October, 2011}

\usepackage{subfigure}
\usepackage{booktabs}
\usepackage{natbib}

@@

One of the morphotactic challenges met in defining a finite-state transducer for Kyrgyz is that many finite verb forms have ``irregular'' negative forms.  While the paradigms are completely regular, the negative morphotactics are not regularly derived from the affirmative forms.  There are also several different ``regular'' patterns of alternation.  Listed in table \ref{irregnegs} are a couple examples of a few finite verb paradigms with their negative forms.

@@

In Kyrgyz, there are two basic archiphonemes: a low vowel, \{A\}, and a high vowel, \{I\}.  The high vowel takes the backness and rounding of the preceding vowel, resulting in the values shown in table \ref{vowelharmonyI}.

@@

The back vowel \{A\} also takes the backness and rounding of the preceding vowel, with the exception of when it occurs after /у/---n this case, it has an unrounded variant, as depicted in table \ref{vowelharmonyA}.

@@

Voicing assimilation in Kyrgyz involves the agreement of voicing of two consonants across a syllable boundary.  An example involves the locative and dative suffixes, as shown in table \ref{exvoicing}.

@@

@@

The major work to be done is increasing the size of the lexicon. While a good level of coverage has been achieved with only XX stems, real-word, or production morphological analysers have tens of thousands of stems, even for morphologically-rich languages like Kyrgyz.

Once good coverage has been achieved with a morphological analyser, the next logical step is to start work on morphological and syntactic disambiguation. The average ambiguity of a surface form in the analyser is YY, meaning that there is a lot of work that can be done on disambiguation.

@@

\bibliographystyle{apa}
\bibliography{paper}
%FIXME: "Kazan Tatar ..."

>>>>

\documentclass[a4paper,12pt,onecolumn,oneside]{article}
\usepackage{polyglossia}
\setdefaultlanguage[variant=australian]{english}

@@

\usepackage{natbib}

\usepackage[in]{fullpage}
\usepackage[colorlinks=true,citecolor=black,linkcolor=black,urlcolor=blue]{hyperref}

\usepackage{subfigure}
\usepackage{booktabs}

%\bibpunct{(}{)}{;}{A}{,}{,}
\bibdata{paper}

@@

\title{A finite-state morphological transducer for Kyrgyz}
%\author{Jonathan North Washington, Mirlan Ipasov, and Francis Tyers\\\href{mailto:jonwashi@indiana.edu}{jonwashi@indiana.edu}, \href{mailto:mipasov@gmail.com}{mipasov@gmail.com}, \href{mailto:ftyers@prompsit.com}{ftyers@prompsit.com}}

\author {Jonathan North Washington, \\
Indiana University,\\
Bloomington, IN 47405-7000 (USA)\\
\texttt{jonwashi@indiana.edu} \and 
Mirlan Ipasov,\\
xxx\\
yyy (Kyrgyzstan)\\
\texttt{mipasov@gmail.com} \and
Francis M. Tyers\\
Dept.\ Lleng.\ i Sist. Informàtics \\  
Universitat d'Alacant\\
E-03071 Alacant (Spain)\\
\texttt{ftyers@dlsi.ua.es} 
 }

\date{October, 2011}

@@

One of the morphotactic challenges met in defining a finite-state transducer for Kyrgyz is that many finite verb forms have ``irregular'' negative forms.  While the paradigms are completely regular, the negative morphotactics are not regularly derived from the affirmative forms.  There are also several different ``regular'' patterns of alternation.  Listed in table~\ref{irregnegs} are a couple examples of a few finite verb paradigms with their negative forms.

@@

In Kyrgyz, there are two basic archiphonemes: a low vowel, \{A\}, and a high vowel, \{I\}.  The high vowel takes the backness and rounding of the preceding vowel, resulting in the values shown in table~\ref{vowelharmonyI}.

@@

The back vowel \{A\} also takes the backness and rounding of the preceding vowel, with the exception of when it occurs after /у/---n this case, it has an unrounded variant, as depicted in table~\ref{vowelharmonyA}.

@@

Voicing assimilation in Kyrgyz involves the agreement of voicing of two consonants across a syllable boundary.  An example involves the locative and dative suffixes, as shown in table~\ref{exvoicing}.

@@

\subsection{Number of stems}

Table~\ref{table:stems} gives the approximate number of stems in each main word class. The numbers
are approximate as some words may have two entries for one stem (for example if a set of forms
is irregular).

\begin{table}
\begin{center}
\begin{tabular}{lr}
\hline
Part of speech & Number of stems\\
\hline
Noun & 4,731\\
Adjective & 924\\
Verb & 896 \\
Proper noun & 745\\
Adverb & 213\\
Numeral & 50\\
Postposition & 42\\
Conjunction & 36\\
Pronoun & 29\\
Determiner & 27\\
\hline
Total:     & 7,333\\
\hline
\end{tabular}
\caption{FIXME}
\label{table:stems}
\end{center}
\end{table}

\subsection{Coverage}

To calculate the naïve coverage\footnote{FIXME} of the analyser, two corpora were used,
 the first was a database dump of the Kyrgyz Wikipedia, dated XX-XX-XX,\footnote{DUMP NAME} 
which was processed with the programs YY and ZZ to extract sentences. The second was FIXME.

Both corpora were split into 10 equal parts, and coverage was calculated over each part 
separately in order to calculate the standard deviation of the mean. As can be seen 
from table~\ref{table:cov}, running the analyser on the RFERL gives a higher coverage, 
because the text is more homogenous, and there is less non-Kyrgyz (e.g. Russian) text.

\begin{table}
\begin{center}
\begin{tabular}{lrrrr}
\hline
Corpus           & Tokens    & Known     & Naïve coverage    & Mean ambiguity\\
\hline
Kyrgyz Wikipedia & 329,524   & 269,573   & 81.84 $\pm$ 3.23  & 4.53\\
RFERL Kyrgyzstan & 4,107,555 & 3,586,000 & 87.37 $\pm$ 1.23  & 4.26\\
\hline
\end{tabular}
\caption{FIXME}
\label{table:cov}
\end{center}
\end{table}

The column 'mean ambiguity' gives the average number of analyses for each surface form 
given when analysing this corpus.

@@

The major work to be done is increasing the size of the lexicon. While a good level of coverage has been achieved with only 7,333 stems, real-word, or production morphological analysers have tens of thousands of stems, even for morphologically-rich languages like Kyrgyz.

Once good coverage has been achieved with a morphological analyser, the next logical step is to start work on morphological and syntactic disambiguation. As can be seen from the figures for mean ambiguity, there is a lot of work that can be done on disambiguation.

@@

\bibliographystyle{fullname}
\bibliography{paper}

%FIXME: "Kazan Tatar ..."
